Using Interactive Graphics to Visualize Evening Grosbeak Movement and Migration Patterns

### 2024-05-13 paper.qmd
changes, just need to go through citations! (and image sizing problems)

<p><small><em>Figure 1. Evening Grosbeak range map by American Bird Conservatory</em></small></p>

@@

![Default Shiny App](default_shiny.png){.lightbox}

@@

![Iris dropdown menus Shiny App](iris_dropdown.png){.lightbox gallery = "iris"}
![Dropdown menu](dropdown.png){.lightbox gallery = "iris"}

@@

@@

<p>The interactive Evening Grosbeak platform was made using Shiny. From the navigation bar, the user can choose from "<em>Deployments and Tracking</em>," "<em>Winter Roosting</em>," or "<em>About</em>." Each navigation page includes multiple tabs for exploration.</p>

@@

### <em>Deployments and Tracking</em>
<p>Upon launching the app, the user is met with the <em>Deployments and Tracking</em> page. This page first includes the map of each bird's first deployment.The user can click through the map of different clusters until they find a specific bird at a deployment site that they want to look furhter into. Upon clicking on this circle marker, the app grabs the data for this bird and stores it to use in the next tab.</p>

@@

@@

![Chosen bird map](chosenbird.png)
![Chosen group map](chosengroup.png)

@@

![Winter roosting location map](wintermap.png)
![Winter roosting group map](wintergroups.png)

@@

-   what I have been able to learn and showcase through this project

>>>>

<!--<p><small><em>Figure 1. Evening Grosbeak range map by American Bird Conservatory</em></small></p> -->

@@

![Default Shiny App](default_shiny.png)

@@

![Iris dropdown menus Shiny App](iris_dropdown.png)
![Dropdown menu](dropdown.png)

@@

<em>Leaflet</em> is another package for data visualizations in R. It comes from a JavaScript library for interactive maps. With this package, one can customize maps with different markers, shadings, pop-ups, lines, and more. This is a very popular tool that pairs well with the R software for exploring spatial data.

Below is a basic example of creating a leaflet map.

```{r, echo=TRUE, include=TRUE, error=FALSE, warning=FALSE, message=FALSE, eval = TRUE}
#| code-fold: show
#| code-summary: "Click to show leaflet code"
library(leaflet)

map <- leaflet() %>%
  setView(lat = 44.5956, lng = -75.1691, zoom = 13) %>%  # Setting initial view to London
  addTiles()  # Add default OpenStreetMap tiles

map
```

@@

<p>The interactive Evening Grosbeak platform was made using Shiny in R. From the navigation bar, the user can choose from "<em>Deployments and Tracking</em>," "<em>Winter Roosting</em>," or "<em>About</em>." Each navigation page includes multiple tabs for exploration.</p>

@@

### <em>Deployments and Tracking</em>
<p>Upon launching the app, the user is met with the <em>Deployments and Tracking</em> page. This page first includes the leaflet map of each bird's first deployment.The user can click through the map of different clusters until they find a specific bird at a deployment site that they want to look furhter into. Upon clicking on this circle marker, the app grabs the data for this bird and stores it to use in the next tab.</p>

@@

```{r, eval = FALSE}
#| code-fold: show
#| code-summary: "Click to see output code for this map"
output$map <- renderLeaflet({
      leaflet() %>%
        addProviderTiles(providers$CartoDB.PositronNoLabels) %>% 
        setView(lat = 15, lng = 0, zoom = 1.5)
    })

# Render circle markers
observe({ 
  map <- leafletProxy("map", data = mapDATA())
  map %>% clearMarkers()
  if (!is.null(mapDATA())) {
    map %>% 
      addCircleMarkers(lat = ~tagDepLat, 
                       lng = ~tagDepLon, 
                       layerId = ~motusTagDepID, 
                       clusterOptions =  markerClusterOptions()
      )
  }
})

# Show popup on click
observeEvent(input$map_marker_click, {
  click <- input$map_marker_click
  text<-paste("ID", click$id, "<br>Lattitude ", round(click$lat, 2), "<br>Longtitude ", round(click$lng,2))

  map <- leafletProxy("map")
  map %>% clearPopups() %>%
    addPopups(click$lng, click$lat, text)
})
```

@@

![Chosen bird map](chosenbird.png)

```{r, eval=FALSE}
#| code-fold: show
#| code-summary: "Click to show code for the chosen bird output"

output$Click_plot <- renderLeaflet({
  click <- input$map_marker_click
  if (!is.null(click)){
    bird <- DATA[which(DATA$motusTagDepID == click$id), ]

    # Go through full data set now to check for this bird with matching id
    one_bird <- eg_df %>%
      filter(eg_df$motusTagDepID == bird$motusTagDepID)

    # sort one_bird by timestamp to ensure you are getting very first pinpoint
    one_bird <- one_bird %>%
      arrange(tsCorrected)
    firstDep <- one_bird %>%
      slice(1)

    # firstDep uses deployment coordinates to mark
    # one_bird uses receiver coordinates to mark
    one_bird <- one_bird[-1, ] # take out deployment site (initial point and point of firstDep)
    # # rename lat and long for each data set
    firstDep <- firstDep %>%
      rename(
        lat = tagDepLat,
        lon = tagDepLon,
      )
    one_bird <- one_bird %>%
      rename(
        lat = recvDeployLat,
        lon = recvDeployLon
      )
    # join the data sets (vertically) (first row should be deploymnent site)
    # make copies of datasets with desired data
    firstDepCopy <- firstDep %>%
      select(motusTagDepID, lat, lon, tsCorrected, ts)
    one_birdCopy <- one_bird %>%
      select(motusTagDepID, lat, lon, tsCorrected, ts)
    full_bird <- bind_rows(firstDepCopy, one_birdCopy)

    new_full_bird <- full_bird %>%
      group_by(lat, lon) %>%
      slice(1)

    popupInfo = paste("Lat = ", new_full_bird$lat, ", Lon = ", new_full_bird$lon)

    leaflet() %>%
      addProviderTiles(providers$CartoDB.PositronNoLabels) %>%
      setView(lat = 15, lng = 0, zoom = 1.5) %>%
      addAwesomeMarkers(data = firstDep,
                        lng = ~lon,
                        lat = ~lat) %>%
      addCircleMarkers(data = new_full_bird,
                       lng = ~lon,
                       lat = ~lat,
                       color = "red",
                       popup = popupInfo,
                       clusterOptions = markerClusterOptions()) %>%
      addPolylines(data = new_full_bird,
                   lng = ~lon,
                   lat = ~lat,
                   color ="blue",
                   weight = 1)

  }
})

```

![Chosen group map](chosengroup.png)

@@

![Winter roosting location map](wintermap.png)

![Winter roosting group map](wintergroups.png)

@@

Creating this interactive platform has transformed my data science knowledge and has blended together many of the coursework topics I have spent years learning. I have been able to expand on my previous knowledge of data cleaning and manipulation when working with this spatial data, which is not a type of data I have gotten the chance to work with before. Working with leaflet has broadened my understanding on the possibilities of data visualizations within R. Working with a large dataset effectively has brought me many tools that will be necessary for working with impactful data and making it accessible to others.
